Model for DES Type Ia Supernova Cosmology Analysis

### 2016-05-22 tex/toy.tex
A fixed up version that I think is rigorous

The ambition is to construct a model to determine the likelihood of DES-like
supernova data. 
The conceptual structure of the model is shown in Figure~\ref{pgm:fig}. 
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=6.5in]{../results//hdpgm.pdf} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   \label{pgm:fig}}
\end{figure}

The model has the following types of observables $O$ that
may be observed for each supernova:
\begin{itemize}
\item $c$: Counts of the photometric measurements (light curves).
\item $s$: Counts of the spectroscopic measurements.
\item $z$: Redshift.
\item $T$: Transient type.
\end{itemize}
The measurements made of the observables are labeled  with a subscript ``o''.

@@

The model describes transients and their observables.  We are particularly interested in the pdf of those observables
given the model.  The pdf is dependent on the sample and we consider two such samples: an underlying sample
that represents the distribution of all possible transients and observables accommodated by the model; and
an analysis sample that represents the distribution of transients and observables that end up being analyzed.
The analysis sample differs from the underlying sample in that the transients of the former 
must be discovered, pass photometric selection criteria, and for the SN~Ia sample be selected for spectroscopic
observation, have a successful type from the spectrum, and be typed as SN~Ia.
The pdf of the underlying sample is denoted as $p(O|M,M_I)$ whereas the pdf of the analysis sample, the likelihood,
is denoted $\mathcal{L}_S(O | M, M_I)$. 
The two are related by
\begin{align}
% \mathcal{L}_S(O_o | M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\mathcal{L}(O_o | M, M_I) & = \frac{p_S(O_o|M, M_I)}{\int  p_S(O|M, M_I)dO},
\label{likelihood:eqn}
\end{align}
where
\begin{align}
p_S(O|M, M_I) &= p(O, disc=True, anal=True, sobs=True, type=True, T=SNIa |M, M_I).
\end{align}

@@

We now consider the terms in the likelihood contributed by each transient and omit the $\alpha$ superscript. 
The dependencies of the parameters are charted through the PGM to give for transients with a  spectroscopic type measurement
\begin{align*}
p(O| M, M_I) & = P(T |\tau) p(z | M) p(c | L, M) p (s|L, M),
\end{align*}
and for transients without a type measurement (not relevant for the SN~Ia-only analysis but included for completeness)
\begin{align*}
p(O| M, M_I) & = p(z | M) p(c | L, M),
\end{align*}
where the photometric and spectroscopic observations have independent measurement uncertainties.

The probability density for an object to enter the analysis sample can be broken out
\begin{align}
p_S(O|M, M_I) &= p(O, disc=True, anal=True, sobs=True, type=True, T=SNIa |M, M_I)\\
&= P(T, T=SNIa| s, c, disc=True, anal=True, sobs=True, type=True, M, M_I)\\
& \times  p(s, type=True | c, disc=True, anal=True, sobs=True, M, M_I)\\
& \times  p(c, disc=True, anal=True, sobs=True| M, M_I)\\
&= P(T, T=SNIa| M, M_I)  p(s, type=True | M, M_I) \\
& \times  p(c, disc=True, anal=True, sobs=True| M, M_I).
\end{align}

@@

\section{Probabilities: Minding Ones P's and p's}
Section~\ref{likelihood:sec} introduces the probabilities and probability distribution functions that describe the likelihood
and posterior for the model shown in Figure~\ref{pgm:fig}.
In this Section we present one possible model by specifying those probabilities; the assembly of other models should
resemble this example.

@@

The type of the object $\tau$ is drawn from the Bernoulli distribution 
\begin{equation}
\tau | p_{SNIa} \sim \mathcal{B}(1-p_{SNIa}),
\end{equation}
with the convention that $T=0$ corresponds to SN~Ia.

@@

\item  The luminosity distance comes from the physics-theory function
\begin{equation}
d_L = d_L(z; \Omega_M, w0),
\end{equation}
the flux for an observation at date $t$ is
\begin{equation}
f = 10^{-0.25 \left(M_{SNIa} - \alpha x_1 + \beta c \right)}\frac{1}{4\pi d_L^2} F_{SALT2}((t-t_0)/(1+z); x_0,  x_1, c, \lambda/(1+z)),
\end{equation}
and expected counts is
\begin{equation}
c_\star = 10^{Z/2.5}f.
\end{equation}
The distribution of counts is 
\begin{align*}
c | c_\star &\sim \mathcal{N}(c_\star, C_c)\\
s | c_\star &\sim \mathcal{N}(c_\star, C_s),
\end{align*}
where $C_c$ and $C_s$ are the measurement covariances for the photometric and spectroscopic observations respectively.

@@

\subsection{$p_S(O|M, M_I)$: Sample Selection Probability}
We have listed the probabilities for the underlying distribution.  We now describe the probabilities for making it into the analysis sample.
There are three terms.

\begin{itemize}
\item $P(T, T=SNIa| M, M_I)$: the probability a type measurement $T$ yields an SN~Ia type.  It depends
on rate parameters as passed through in Eqn.~\ref{typepdf:eqn}.
\item $p(s, type=True | M, M_I)$: the probability of a spectral measurement $s$ that yields a successful typing.
\item $p(c, disc=True, anal=True, sobs=True| M, M_I)$
\end{itemize}
%\subsection{$\epsilon(O, M, M_I)$: Sample Inclusion Efficiency}
%The probability that a transient becomes a part of the analysis sample is given by  \begin{equation}
%P(typed, spec, sample, discovered | s, c, M, M_I) = \epsilon_T(s, M, M_I) \epsilon_O(c) \epsilon_S(c) \epsilon_D(c)
%\end{equation}
%and is dependent
%on the properties of the experiment and choices made by the experimenter.
%The probability of typing depends on the spectroscopic data whereas the probability of spectroscopic observation,
%sample selection, and discovery depend on photometry.
%
%\begin{itemize}
%\item The probability the transient is discovered 
%in a single image depends on the transient signal-to-noise $\epsilon(STON)$.  For
%an image with noise $N$ the efficiency for discovery is $\epsilon(c/N)$, and the probability of discovery in at least one image
%\begin{equation}
%\epsilon_D(c) = 1-\prod_{\beta \in images} \left(1-\epsilon(c_\beta/N_\beta)\right).
% \end{equation}
%\item
%As an illustrative example we write out the probability of a sample selection requiring two photometric points above
%a signal-to-noise threshold $STON_T$:
%\begin{align*}
%\epsilon_S(c) = &1 - \prod_{\beta \in images} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\mathcal{N}(c_\star-c_\beta, C_c)\\
%&  - \sum_{\alpha \in images}\mathcal{H}(c_\alpha/N_\alpha; STON_T) \mathcal{N}(c_\star-c_\alpha, C_c)\prod_{\beta \in \{images-\alpha\}} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\mathcal{N}(c_\star-c_\beta, C_c),
% \end{align*}
% where $\mathcal{H}$ is the Heaviside step function (other sample selection cases are similar but with different combinatorics).
%\item The probability of obtaining a spectroscopic observation of a transient is given as $\epsilon_O(c)$.  Each follow-up telescope
%has its own criteria: OzDES is magnitude limited, VLT prefers transients on faint hosts, others may try to get a random sample uniform in
%redshift.  Mat Smith and Chris D'Andrea should be able to provide this term.
%\item The probability that the spectroscopic observation yields a type is $\epsilon_T( s, M, M_I)$.  This is a function we have to
%figure out!
%\end{itemize}

\subsection{$\int  p_S(O|M, M_I)dO$: Normalizing the Likelihood}
We could write
\begin{align*}
P(sample, discovered |c) & =P(sample| discovered, c)P(discovered |c)\\
&= \epsilon_S(c)
\end{align*}
if the sample selection criteria are selected such that $P(discovered | sample, c)=1$.
This would have significant
computational benefits in the evaluation of the denominator of Eqn.~\ref{likelihood:eqn}.

>>>>

The ambition is to construct a model that predicts the observations of
transients.
The conceptual structure of the model is shown in Figure~\ref{pgm:fig}. 
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=6.5in]{../results//hdpgm.pdf} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   \label{pgm:fig}}
\end{figure}

@@

The model has the following types of observables $O$ that
may be observed for each transient:
\begin{itemize}
\item $c$: Counts of the photometric measurements (light curves).
\item $s$: Counts of the spectroscopic measurements.
\item $z$: Redshift.
\item $T$: Transient type.
\end{itemize}
The measurements made of the observables are labeled  with a subscript ``o''.

We are interested in what the model predicts for the pdf of those observables.  The pdf is dependent on the population and we consider two such populations.  The first is
the underlying population
of all possible transients and observables accommodated by the model: the pdf for this
distribution is $p(O|M,M_I)$  and in this note unsubscripted $p$ and $P$'s are probabilities
for this popiulation.
The second is
the analysis sample that represents the population of transients and observables that end up being analyzed.
The analysis population differs from the underlying population in that the transients of the former 
must be discovered, pass photometric selection criteria, and for the SN~Ia sample be selected for spectroscopic
observation, have a successful type from the spectrum, and be typed as SN~Ia.
The $p$ and $P$ probabilities for the analysis population are subscripted with $S$.
For the SN~Ia sample the two are related by
\begin{align}
% \mathcal{L}_S(O_o | M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\mathcal{L}_S(O_o | M, M_I) & = \frac{p_S(O_o|M, M_I)}{\int  p_S(O|M, M_I)dO},
\label{likelihood:eqn}
\end{align}
where
\begin{align}
p_S(O|M, M_I) &= p(O, disc=True, anal=True, sobs=True, type=True, T=SNIa |M, M_I)\\
&= p(O, disc=True, anal=True, sobs=True, type=True| T=SNIa, M, M_I)\\
& \times P(T=SNIa| M, M_I)\\
&= p(T, z, s, type=True| T=SNIa ,M, M_I)\\
& \times p(c, disc=True, anal=True, sobs=True| T=SNIa,  M, M_I)\\
& \times P(T=SNIa| M, M_I) \\
&= p(T, Œz, type=True|s,  T=SNIa, M, M_I) \\
& \times P(disc=True, anal=True, sobs=True| c, T=SNIa, M, M_I) \\
& \times p(s| T=SNIa, M, M_I) p(c| T=SNIa, M, M_I) P(T=SNIa| M, M_I).
\end{align}
We anticipate that the discovery, photometric selection, spectroscopic
targeting, and photometric measurements are independent of the typing success, and the
spectral, redshift and typing measurements.

@@

%We now consider the terms in the likelihood contributed by each transient and omit the $\alpha$ superscript. 
%The dependencies of the parameters are charted through the PGM to give for transients with a  spectroscopic type measurement
%\begin{align*}
%p(O| M, M_I) & = P(T |\tau) p(z | M) p(c | L, M) p (s|L, M),
%\end{align*}
%and for transients without a type measurement (not relevant for the SN~Ia-only analysis but included for completeness)
%\begin{align*}
%p(O| M, M_I) & = p(z | M) p(c | L, M),
%\end{align*}
%where the photometric and spectroscopic observations have independent measurement uncertainties.
%

@@

\section{Probabilities: Minding Ones P's and p's}
Section~\ref{likelihood:sec} introduces the probabilities and probability distribution functions that describe the likelihood
and posterior for the model shown in Figure~\ref{pgm:fig}.
In this Section we present one possible model by specifying underlying the
probabilities; the assembly of other models should
resemble this example.

@@

The type of the object $\tau$ is drawn from the Bernoulli distribution 
\begin{equation}
\tau | p_{SNIa} \sim \mathcal{B}(1-p_{SNIa}),
\end{equation}
with the convention that $T=0$ corresponds to SN~Ia.
This is a term of $p_S(O|M, M_I)$

@@

\item  The luminosity distance comes from the physics-theory function
\begin{equation}
d_L = d_L(z; \Omega_M, w0),
\end{equation}
the flux for an observation at date $t$ is
\begin{equation}
f = 10^{-0.25 \left(M_{SNIa} - \alpha x_1 + \beta c \right)}\frac{1}{4\pi d_L^2} F_{SALT2}((t-t_0)/(1+z); x_0,  x_1, c, \lambda/(1+z)),
\end{equation}
and expected counts is
\begin{equation}
c_\star = 10^{Z/2.5}f.
\end{equation}
The distributions of counts are 
\begin{align*}
c | c_\star &\sim \mathcal{N}(c_\star, C_c)\\
s | c_\star &\sim \mathcal{N}(c_\star, \sigma_s),
\end{align*}
where $C_c$ and $C_s$ are the measurement covariances for the photometric and spectroscopic observations respectively.  These are terms of $p_S(O|M, M_I)$.

@@

\subsection{$p_S(O|M, M_I)$: Sample Selection Probability}
We have listed the probabilities for the underlying distribution.  We now describe the probabilities 
of the two terms that involve sample selection.

\begin{itemize}
\item Spectroscopic term: $p(T, z, type=True|s,  T=SNIa, M, M_I) $. 
This term can be reduced as
\begin{align*}
p(T, z, type=True|  s,  T=SNIa, M, M_I) &= p(T|  s,  T=SNIa, M, M_I) \\
&\times P(type=True| s,  T=SNIa, M, M_I)\\
&\times p(z|s,  T=SNIa, M, M_I),
 \end{align*}
with the (reasonable) approximation that the three measurements are
independent.

Condition §$type=True$: The probability of successful spectroscopic typing needs to be determined.
A simple possibility is that  it only depends on the signal-to-noise of the spectroscopic observation
$\epsilon_T(ston)$.  Then given a  spectroscopic noise of $\sigma_s$
\begin{equation}
P(type=True| s,  T=SNIa, M, M_I) = \epsilon_T(s/\sigma_s).
\end{equation}

Assembling all the terms gives
\begin{align*}
p(T, z, type=True|  s,  T=SNIa, M, M_I) &= \delta_{T, SNIa} \delta(z-z_{host}) \epsilon_T(s/\sigma_s) .
 \end{align*}

\item Photometric term: $P(disc=True, anal=True, sobs=True| c, T=SNIa, M, M_I)$.
\begin{align*}
P(disc=True, anal  =True, sobs=True| c, T=SNIa, M, M_I) & =P(anal=True| c, T=SNIa, M, M_I) \\
& \times P(sobs=True|  c, T=SNIa, M, M_I) \\
& \times P(disc=True| c, T=SNIa, M, M_I),
 \end{align*}
taking that the random processes that determine discovery, sample selection, and
spectroscopic observation are independent.

Condition $disc=True$:
The probability for discovery of a transient in a single image depends on the
signal-to-noise $\epsilon_D(ston)$.  The probability of discovery given all
the images is
\begin{equation}
P(disc=True | c, T=SNIa, M, M_I)
=  1-\prod_{\beta \in images} \left(1-\epsilon_D(c_\beta/N_\beta)\right).
\end{equation}

Condition $anal=True$:
As an illustrative example we write out the probability of a analysis-sample selection requiring two photometric points above
a signal-to-noise threshold $STON_T$: other sample selection cases are similar but with different combinatorics
\begin{align*}
P(anal=True  | c, T=SNIa, M, M_I)= &1 - \prod_{\beta \in images} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\\
&  - \sum_{\alpha \in images}\mathcal{H}(c_\alpha/N_\alpha; STON_T) \prod_{\beta \in \{images-\alpha\}} (1-\mathcal{H}(c_\beta/N_\beta; STON_T)),
 \end{align*}
 where $\mathcal{H}$ is the Heaviside step function.

Condition $sobs=True$:
The probability a transient has a spectroscopic observation.
Mat Smith and Chris D'Andrea should be able to provide this term.
Until we get this, lets make the simplifying ansatz that
\begin{align*}
P(sobs=True | c, T=SNIa, M, M_I)= & const.
 \end{align*}
